BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2024-08-16 manuscript/manuscript_APPS.rmd
data

The label generation functionality is provided explicitly by two other programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
The office suite costs money, and in our experience is notoriously finicky, further it's functionality ends with label creation; while we expect this tools exists in open source alternatives such as Libreoffice and OpenOffice, we expect them to have similar issues with use. 
PLabel has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface functionality, unfortunately it does not include data cleaning functionality; while some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows emulators like Wine. 
The popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections. 
However, not all herbaria use Symbiota and many have original database systems which they maintain. 
Furthermore, many collectors like to process their own labels, especially as they are likely to be sending sets to different institutions. 
Accordingly, Symbiota's functionality should exist in an ecosystem with alternative systems.

@@

## Usage
All steps of BarnebyLives except for label generation are run from within Rstudio. 
Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over 100 real-world examples are on a Google Sheets accessible from the page.
BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
Virtually all of the on-disk memory associated with these data are for storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation. 
Functions which require the on-disk data require a path to the data as an argument.  
Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.  
We anticipate most personal BarnebyLives instances will be less than several gigabytes (ours covering all of the conterminous Western U.S. is XX GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g. pdflatex, lualatex, xelatex), and the open source command line tools pdfjam and pdftk.
While these steps are run through bash, we have wrapped them in R functions which bypass the need to enter the commands to a terminal.
Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.

@@

Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use. 
While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Typo', na.rm = T)` typos, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances of an outdated circumscription applied. 
At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error.

@@

While numerous tools have been developed for cleaning of existing herbarium and museum records, few help with ensuring that the entered data are accurate (@patten2024geographic),

Accessioning often times relies on the use of the Office Suite of programs, and may utilize other costly softwares, such as ArcPro, or Adobe Acrobat. 
While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet formats

LaTeX offers well documented and detailed functionality for customizing labels for individuals and institutions... 
While the program

>>>>

The label generation functionality is provided explicitly by two other programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
The office suite costs money, and in our experience is finicky, further it's functionality ends with label creation; while we expect this tools exists in open source alternatives such as Libreoffice and OpenOffice, we expect them to have similar issues with use. 
PLabel has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface functionality, unfortunately it does not include data cleaning functionality; while some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows emulators like Wine. 
The popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections.  
Symbiota offers functionality similar to our entire 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries' as well (see FIG). 
However, not all herbaria use Symbiota and many have original database systems which they maintain. 
Furthermore, many collectors like to generate their own labels, especially as they are likely to be sending different sets of collections to different institutions. 
Accordingly, Symbiota's functionality should exist in an ecosystem with alternative systems. 
In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats
after utilizing BLs data cleaning utilities.

@@

## Usage
All steps of BarnebyLives except for label generation are run from within Rstudio. 
Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over 100 real-world examples are on a Google Sheets accessible from the page.
BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
Virtually all of the on-disk memory associated with these data are for storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation. 
Functions which require the on-disk data require a path to the data as an argument.  
Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.  
We anticipate most personal BarnebyLives instances will be less than several gigabytes (ours covering all of the conterminous Western U.S. is XX GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g. pdflatex, lualatex, xelatex), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk]().
While these steps are run through bash, we have wrapped them in R functions which bypass the need to enter the commands to a terminal.
Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.

@@

Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use. 
While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Typo', na.rm = T)` typos, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances where an outdated circumscription was applied. 
At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error.

@@

While numerous tools have been developed for cleaning of existing herbarium and museum records, few help with ensuring that the entered data are accurate (@patten2024geographic). 
By utilizing both R and LaTeX, and having publicly available source code on Github, this program allows most users immediate familiarity with the system for troubleshooting issues, and implementing upgrades and modifications on project branches.

Accessioning often times relies on the use of the Office Suite of programs, and may utilize other costly software, such as ArcPro, or Adobe Acrobat. 
While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet formats.

LaTeX offers well documented and detailed functionality for customizing labels for individuals and institutions. 
Very good documentation of LaTeX capabilities are offered in multiple areas for instance via the [Overleaf](https://www.overleaf.com/learn)) project. 
While the program
